Pre-/Post-Trikafta in CFRD

### 2022-02-02 Christine Chan/Pre-post Trikafta/pre_post_trikafta_analysis.Rmd
Update pre_post_trikafta_analysis.Rmd

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(lmerTest)
library(broom.mixed)
library(performance)
library(lubridate)
library(plotly)
library(knitr)
knitr::opts_chunk$set(echo = FALSE,fig.height = 8,fig.width = 10)
if(Sys.info()["sysname"] == "Windows"){
  home_dir = "S:/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
knitr::opts_knit$set(root.dir = home_dir)
```

@@

```{r}
data = df
outcome = "HbA1c"
# Piecewise model function
piecewise = function(outcome,data = df){
  # Add spline column
  data$Change = ifelse(data$Time < 0,0,1)
  data$Change = data$Change * data$Time
  # Model
  f = as.formula(paste0(outcome,"~Time + Change + (1|CFF.ID)"))
  mod = lmer(f,data = data)
  res = tidy(mod,"fixed")
  res$effect = NULL
  # Coordinates for model plotting
  x1_start = -365
  y1_start = res$estimate[1] + res$estimate[2]*x1_start
  x1_end = 0
  y1_end = res$estimate[1] 
  x2_start = 0
  y2_start = res$estimate[1] 
  x2_end = 365
  y2_end = res$estimate[1] + (res$estimate[2]+res$estimate[3])*x2_end
  # Plot
  p = ggplot(data,aes_string(x = "Time",y = outcome,group = "CFF.ID",color = "Site")) +
    geom_point(alpha = 0.5) + geom_line(alpha = 0.2) +
    annotate("segment",x=x1_start,y=y1_start,xend=x1_end,yend=y1_end,color = "blue",size = 1) +
    annotate("segment",x=x2_start,y=y2_start,xend=x2_end,yend=y2_end,color = "blue",size = 1) +
    xlab("Days from Trikafta Start") + theme_bw()
  # Interactive plot and model results
  return(list(plot = ggplotly(p),table = res))
}
```

@@

@@

```{r results='asis'}
a1c = piecewise(outcome = "HbA1c",df)
ggplotly(a1c$plot)
kable(a1c$table,digits = 3)
```

>>>>

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(lmerTest)
library(nlme)
library(broom.mixed)
library(segmented)
library(performance)
library(lubridate)
library(plotly)
library(knitr)
knitr::opts_chunk$set(echo = FALSE,fig.height = 8,fig.width = 10)
if(Sys.info()["sysname"] == "Windows"){
  home_dir = "S:/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
knitr::opts_knit$set(root.dir = home_dir)
```

@@

```{r}
data = df
outcome = "HbA1c"
# Piecewise model function
piecewise = function(outcome,data = df){
  # Add spline column
  data$Change = ifelse(data$Time < 0,0,1)
  data$Change = data$Change * data$Time
  # Model
  f = as.formula(paste0(outcome,"~Time + Change + (1|CFF.ID)"))
  mod = lmer(f,data = data)
  res = tidy(mod,"fixed")
  res$effect = NULL
  res$term[2] = "Slope Prior"
  # Segmented fit
  fit = lme(as.formula(paste0(outcome,"~Time + Change")),random = ~1|CFF.ID,data = data,na.action = na.omit)
  seg = suppressWarnings(segmented(fit, seg.Z = ~Time, npsi=1))
  # Model check if necessary
  check = check_model(mod)
  # Coordinates for model plotting
  x1_start = -365
  y1_start = res$estimate[1] + res$estimate[2]*x1_start
  x1_end = 0
  y1_end = res$estimate[1] 
  x2_start = 0
  y2_start = res$estimate[1] 
  x2_end = 365
  y2_end = res$estimate[1] + (res$estimate[2]+res$estimate[3])*x2_end
  # Plot
  p = ggplot(data,aes_string(x = "Time",y = outcome,group = "CFF.ID",color = "Site")) +
    geom_point(alpha = 0.5) + geom_line(alpha = 0.2) +
    annotate("segment",x=x1_start,y=y1_start,xend=x1_end,yend=y1_end,color = "blue",size = 1) +
    annotate("segment",x=x2_start,y=y2_start,xend=x2_end,yend=y2_end,color = "blue",size = 1) +
    xlab("Days from Trikafta Start") + theme_bw()
  # Interactive plot and model results
  return(list(plot = ggplotly(p),table = res,check = check,seg = seg))
}
```

@@

For each outcome, we fit a piecewise linear mixed effects model with random intercept for each person and a "knot" at 0. This approach can be thought of as fitting two separate linear models, while ensuring that the two regression lines meet at Trikafta start. In the model results, "Slope Prior" indicates the slope prior to start of treatment (i.e. the average change in outcome for each "1 unit increase" in time). The "Change" term indicates the *change in slope* after medication start. To find the slope post-Trikafta, you add the "Slope Prior" and "Change" estimates.

@@

```{r results='asis',warning=FALSE}
a1c = piecewise(outcome = "HbA1c",df)
ggplotly(a1c$plot)
kable(a1c$table,digits = 3)
a1c$check
```

`r if(class(a1c$seg)=="lme"){"Automated R package was not able to find a better breakpoint."}`
